Galaxy Zoo: Morphological Classifications for Galaxies in HST Legacy Imaging

### 2016-04-29 writeup/gz_hubble_data.tex
Finished first draft of external comparison section

\subsection{Comparison to other catalogs}\ref{ssec:comparisons}

Compare GZH data to:

\begin{itemize}
    \item \citet[ZEST;][]{sca07} (COSMOS)
    \item Tasca (COSMOS)
    \item Cassata (COSMOS)
    \item Zajmoski (COSMOS)
    \item \sout{GEMS morphologies} \citep{hau07}
    \item \sout{AEGIS morphologies} \citep{lot08}
    \item \sout{GOODS N/S morphologies} \citep{bun05}
\end{itemize}

{\note \textit{Address trends seen in broad morphological classes, possible reasons for difference. Also should attempt to map between the GZH vote fractions and whatever classification systems are used in the above systems.}}

\subsubsection{GEMS}

Morphologies for galaxies in GEMS imaging have been measured by \citet{hau07}, who perform a careful comparison of single-component \sersic{} fits to the F850LP imaging using both the \galfit{} and \gimtwod{} codes. We compare the GZH morphological vote fractions to the \galfit{} parameters, which are evaluated by \citet{hau07} as more reliable due to \galfit's ability to fit multiple galaxies and handle crowded fields.

A single \sersic{} fit is often a poor fit to the light profile of a resolved galaxy, since large numbers of classical spirals have both a de~Vaucouleurs bulge at the center and an exponential disc at larger radii. If a single parameter is used, however, a common division in the literature is to split on the \sersic~index $n$ into disc-dominated galaxies ($n<2.5$) and bulge-dominated galaxies ($n>2.5$), where $n$ is the exponent in the radial surface brightness profile such that $\Sigma(r)=\Sigma_e \exp{\left(-\kappa[(r/r_e)^{1/n} - 1]\right)}$.

A 1\arcsec~positional match for the GEMS galaxies in Table~9 in \citet{hau07} to the GZH sample gives 8,846~galaxies found in both samples. We analyze galaxies for which the \galfit{} parameters are flagged as being constrained within plausible physical limits, indicating a moderately robust fit. There is a strong correlation between $n$ and \pbest; the average value for disk-dominated galaxies ($n<2.5$) is $\langle$\pbest$\rangle=0.44\pm0.26$, compared to $\langle$\pbest$\rangle=0.30\pm0.25$ for bulge-dominated galaxies ($n>2.5$). The correlation is stronger for bulge-dominated galaxies; for $n>2.5$, most galaxies have \pbest$<0.2$. In contrast, while $n<2.5$ do have higher \pbest~values, the spread is larger and there are significant numbers of galaxies even at the highest \pbest{} values.

Inverting the comparison, we split the GZH galaxies into three samples: ``clean'' disks (\pbest$\ge0.8$), ``clean'' ellipticals (\pbest$\le0.2$), and ``intermediate'' ($0.2<$\pbest$<0.8$) galaxies \citep[eg,][]{sch14}. Splitting by $n<2.5/n>2.5$, clean disks are $85\%/17\%$, intermediates $84\%/16\%$, and clean ellipticals $59\%/41\%$. There is a strong dependence for these ratios on the brightness of the galaxy, which affects both automatic fits and visual accuracy. Limiting the sample to galaxies with $m_{850LP}<21$ gives fractions of: clean disks are $77\%/23\%$, intermediates $52\%/48\%$, and clean ellipticals $38\%/62\%$. While the clean disk fraction is largely unchanged, intermediates are more evenly balanced and there is a larger population of smooth galaxies with high $n$, as expected.

We visually inspect galaxies for which the \galfit{} and GZH measurements indicate strong disagreement. Of the 84~galaxies with $n>2.5$ and \pbest$\ge0.80$, almost all are spirals with obvious disks and clearly marked spiral arms. These appear to be cases where \galfit{} fits the bulge at the center, but misses extended emission from the disk. We also examine the 35~galaxies with $n<2.5$ and \pbest$\le0.20$; these galaxies generally have smoother light profiles with no spiral arms, although several have bars and/or the sharper truncation of light near the edges that indicates an S0 galaxy. These galaxies are also significantly bluer $(\langle m_{606}-m_{850}\rangle=0.27\pm3.3)$ compared to the majority of galaxies classified as ``smooth'' in GZH $(\langle m_{606}-m_{850}\rangle=1.2\pm1.7)$. We interpret this as evidence for the fact that single-component radial fits are often inadequate to accurately characterize galaxy morphology, and that a simple cut risks significant amounts of misclassified objects.

Clumpy galaxies in GZH have \galfit{} parameters more consistent with disky, $n<2.5$ galaxies than spheroids (Figure~\ref{fig:clumpy_gems}). Using the thresholds in Table~\ref{tbl:thresholds}, GZH clumpy galaxies have average values of $n=1.6\pm1.2$ and $R_e=32\pm16$~pix, with only 11\% of clumpy galaxies falling in the bulge-dominated region of $n>2.5$.

\begin{figure}
\begin{center}
\includegraphics[width=0.50\textwidth]{figures/gems_clumpy.png}
\caption{Parameter distributions for GEMS galaxies in \citet{hau07} and GZH. Red and blue points indicate galaxies split by their \sersic~index (at $n=2.5$) into disk- and bulge-dominated galaxies, respectively. Galaxies identified as ``clumpy'' in GZH are labeled in green.}
\label{fig:clumpy_gems}
\end{center}
\end{figure}

\subsubsection{AEGIS}

Galaxies in AEGIS were morphologically classified using non-parametric measurements by \citet{lot08}. They use the Gini coefficient ($G$), which measures the relative inequality in pixel brightness, and $M_{20}$, the second-order moment of the brightest 20\% of the light \citep{lot04}. A linear combination of $G$ and $M_{20}$ to delineate three broad categories of galaxy morphology: E/S0/Sa (bulge-dominated/spheroids), Sb/Sc/Ir (disk-dominated/irregulars), and mergers.

We compare the GZH vote fractions to the morphological categories defined by their $G$-$M_{20}$ values, which are measured in both $V$- and $I$-band. A 1\arcsec{} positional match on AEGIS from \citet{lot08} and GZH yields 7,508~galaxies; of those, we analyze a subsample of 4,031 with reliably-measured morphologies and $S/N>3$ in both bands.

The GZH morphologies are reasonably correlated with the $G$-$M_{20}$ morphological categories. For galaxies labeled as Sb/Sc/Ir by \citet{lot08}, 58\% of the galaxies in $I$-band (61\% in $V$-band) have \pbest{} above a nominal cut of 0.5. E/S0/Sa galaxies have a tighter relation, with 78\% of galaxies in $I$-band (79\% in $V$-band) having $p_\mathrm{smooth}>0.5$. Galaxies with high \pbest{} are largely concentrated in the E/S0/Sa region of $G$-$M_{20}$ space; the small overlaps into mergers and E/S0/Sa are close to what we expect for loosely defined boundaries. Galaxies with low \pbest{} have a much broader range of $G$-$M_{20}$ values, with nearly half lying well within the Sb/Sc/Ir region.

Of the 47~galaxies identified by GZH as mergers in the cross-matched AEGIS sample (selected as \pmerger$>0.5$ and $N_\mathrm{merger}\ge20$), roughly two-thirds fall in the merger region of $G$-$M_{20}$ space defined in \citet{lot08}; almost all the remainder would be classified as Sb/Sc/Ir. Galaxies identified as mergers by $G$-$M_{20}$, in contrast, typically have lower \pmerger{} values; 66\%/50\% in $V$-band and $I$-band, respectively. While the total number of mergers identified in GZH is roughly $\times10$ fewer than the volume-limited $G$-$M_{20}$ sample, we do confirm the result of a non-evolving merger fraction between $0.2<z<1.2$. %The average bulge statistic for the GZH mergers \citep[as defined in][]{sny15} is $F=-0.3\pm0.5$ in $V$-band and $F=-0.2\pm0.5$ in $I$-band.

\subsubsection{GOODS}

Galaxies in the GOODS-N and GOODS-S fields down to a limit of $z_{AB}=22.5$ have been visually classified by a single expert (R.S.~Ellis), inspecting both $z$-band and composite $Viz$ color images \citep{bun05}. These morphologies are assigned a numerical value based on categories in \citet{bri98a}; these are combined into broader categories of ``E/S0'' (classes 0,1,2), ``Spirals'' (classes 3,4,5), and ``Peculiar/Irregular'' (classes 6,7,8). \citet{bun05} use these morphologies to show that there are an increasing proportion of early-type galaxies observed at later times.

Matching the \citet{bun05} catalog to GZH with a 0.5\arcsec{} radius yields 2435 galaxies (1300 in GOODS-N, 1135 in GOODS-S). We compare their broad morphological categories to \pbest{} in GZH; demographics of the separate GOODS-N and GOODS-S samples are highly consistent. ``E/S0'' galaxies are dominated by high smooth fractions in GZH, with an average value of $\langle p_\mathrm{smooth,best}\rangle=0.54\pm0.20$. The effect is stronger when examining sub-divisions; pure E galaxies have the highest $p_\mathrm{smooth,best}$ values, while S0 have the lowest. Visual inspection of galaxies designated as S0 show numerous examples of both edge-on galaxies and visible substructure in the disk, including spiral arms. This is consistent with the design of the GZH decision tree, and likely reflects that the expert categories are using different criteria.

``Spiral'' galaxies in \citet{bun05} have robustly high GZH vote fractions of features and/or disks, with $\langle p_\mathrm{features,best}\rangle=0.59\pm0.06$. The vote fraction increases with Hubble type, with Scd galaxies having the highest average \pbest{} values. ``Peculiar/Irregular'' galaxies have similar ranges of \pbest{} to the ``Spiral'' galaxies, but can be sub-selected with cutoffs on $p_\mathrm{odd}$ and $p_\mathrm{irregular}$ or $p_\mathrm{merger}$. We note that mergers are not strongly correlated in GZH and the expert visual classifications; between 20\%--30\% have GZH merger fractions of zero, while even clear tidal tails can be as low as $p_\mathrm{merger}\sim0.1$. The total number of major mergers is small enough that expert visual inspection can be practically done on all candidates; a combination of methods may be the most accurate way of identifying these rarer objects, especially when the signs of interaction are relatively subtle.

>>>>

\subsection{Comparing GZH morphologies to other catalogs}\label{ssec:comparisons}

All of the Legacy surveys included in the GZH imaging have had morphological catalogs previously published; these catalogs have significant differences in the number of galaxies, size and magnitude limits, classification scheme, and the methods used for measuring morphology. We cross-match several of these catalogs to GZH and compare the results; this is not presented as an endorsement of any particular method, but as an exploration of the strengths and weaknesses of the GZ crowdsourced catalogs as compared to products made with machine learning, automatic fits, and expert visual classification.

The types and accuracies of morphological classification strongly depend on the sample and methods being used. In an attempt to make a consistent comparison between different techniques, we broadly group galaxies classified into three categories: bulge-dominated/elliptical/smooth, disk-dominated/spiral, and irregular/clumpy. These categories are compared to two parameters in GZH: \pbest, which is designed to identify smooth (elliptical) galaxies, and \podd, which is designed to identify deviations from well-formed spirals or S0s and which constitutes a ``catch-all'' for the variety of asymmetric morphologies that can constitute an irregular galaxy.

Morphologies for GEMS galaxies were measured by \citet{hau07}, who use single-component \sersic{} fits to the F850LP imaging. We use parameters from the \galfit{} code, which \citet{hau07} evaluated as more reliable than \gimtwod{} for GEMS due to the ability to fit multiple galaxies in crowded fields. A 1\arcsec~positional match for the GEMS galaxies in Table~9 in \citet{hau07} to the GZH sample gives 8,846~galaxies in both \citet{hau07} and GZH. The main morphological parameter in the automated catalog is the \sersic{} index $n$ defining the radial surface brightness profile. We designate ``elliptical'' galaxies as those with $n>2.5$ and ``spirals''  with $n>2.5$. There is no automatic measurement of irregular or clumpy structure in this catalog.

AEGIS galaxies were morphologically classified using non-parametric measurements by \citet{lot08}. This method uses a combination of the Gini coefficient ($G$), which measures the relative inequality in pixel brightness, and $M_{20}$, the second-order moment of the brightest 20\% of the light \citep{lot04}. A linear combination of $G$ and $M_{20}$ delineates their three broad categories of galaxy morphology: E/S0/Sa (``elliptical''), Sb/Sc/Ir (``spiral''), and mergers (``irregular''). A 1\arcsec{} positional match on AEGIS and GZH yields 4,031 galaxies with reliably-measured morphologies and $S/N>3$ in both $V$- and $I$-bands.

Galaxies in both of the GOODS fields down to a limit of $z_{AB}=22.5$ were visually classified by a single expert (R.S.~Ellis), inspecting both $z$-band and composite $Viz$ color images \citep{bun05}. These morphologies are assigned a numerical value based on categories in \citet{bri98a}; the corresponding morphologies used are ``elliptical'' (classes 0,1,2), ``spirals'' (classes 3,4,5), and ``irregular'' (classes 6,7,8). A 0.5\arcsec{} matched radius yields 2,435 galaxies (1300 in GOODS-N, 1135 in GOODS-S) in both \citet{bun05} and GZH.

COSMOS galaxies have multiple published datasets automatically classifying their morphologies, all using a variation of non-parametric measurements. \citet{cas07} use a combination of concentration ($C$), asymmetry ($A$), $G$, and $M_{20}$ \citep{cas05} to classify all galaxies with $m_\mathrm{I,petro}<25$, and use an empirical division based on several hundred training images to assign galaxies to discrete morphological categories. A similar method is employed by \citet{tas11}, using the same non-parametric indices but with a different method of calculating the Petrosian radius and total light profile. They employ a nearest-neighbors method is used to assign morphological categories for the entire sample based on a small training set. \citet[][ZEST]{sca07} use $C$, $A$, $G$, $M_{20}$, the galaxy ellipticity ($\epsilon$), and \sersic{} index ($n$) to quantify the galaxy shapes; a principal component analysis is used to assign galaxies to discrete morphological categories. The categories in all three methods are split into ellipticals, spirals, and peculiar/irregulars. A 0.5\arcsec{} radius search on GZH matches to 76,241 galaxies in \citet{cas07}, 76,608 in \citet{tas11}, and 77,425 in \citet{sca07}.

\begin{figure*}
\center
\includegraphics[width=1.0\textwidth]{figures/comparisons.png}
\caption{Distributions of morphological parameters for galaxies matched in GZH and other published catalogs, split by survey (AEGIS, COSMOS, GEMS, and GOODS). The first and third columns show the fraction of overall galaxies in the matched samples, split by their published morphologies as a function of the GZH \pbest. The corresponding second and fourth columns show data for the same galaxies as a function of \podd. Confidence intervals for each binned fraction in the bin are calculated for a binomial population \citep{cam11}.}
\label{fig:comparisons}
\end{figure*}

Figure~\ref{fig:comparisons} shows the proportion of galaxies as split by their automated/expert visual morphologies for each of the six catalogs matched to GZH. The relation between GZH and the automated/expert morphologies is shown to vary significantly for different surveys --- this may be due to intrisic differences in the galaxies, in the techniques used to measure morphology, or both. The distribution of \pbest{} for AEGIS galaxies is essentially flat for spirals, showing almost no difference in proportion at the highest and lowest ends. Early-type galaxies are at their highest proportion at \pbest$\sim0$, but constitute only 40\% of the galaxies at most. In contrast, GOODS galaxies at low \pbest{} are dominated by early-types, although spiral and irregular galaxies have comparable sizes as low as \pbest$\sim0.2$. In GEMS galaxies, the impact of irregular or peculiar galaxies is not measured; splitting on $n=2.5$ gives an essentially flat proportion of ellipticals to spirals at \pbest$>0.3$. Visual inspection of images of galaxies where both \pbest{} and $n$ are high show that the majority are obvious spirals but with prominent bulges, indicating that the single-component \sersic{} fit is likely choosing too small of an effective radius and missing the extended disk structure for a large population of galaxies.

Even within the same sample of galaxies in COSMOS, different techniques of measuring morphology show significant differences. Both \citet{cas07} and \citet{tas11} have higher proportions of early-type galaxies at \pbest$\sim0$, but constitute less than half the sample in both cases. The majority of all galaxies in \citet{sca07}, in contrast, are spirals; oddly, the \emph{highest} proportions of spirals are found at the lowest values of \pbest. The decrease in spiral galaxies at higher \pbest{} is almost entirely balanced by an increase in irregulars. Early-types have a small bump near \pbest$\sim0.15$, but never constitute more than 10\% of the sample.

The second set of plots in Figure~\ref{fig:comparisons} indicates that \podd{} is largely an effective method for separating irregular and/or peculiar galaxies from spirals. Despite the somewhat vague wording of the ``anything odd?'' question in GZH, the AEGIS, COSMOS, and GOODS all show a marked increase in the irregular fraction increasing with \podd. Any threshold value for distinguishing between the two, however, depends strongly on survey/redshift~range.

As the data shown in Figure~\ref{fig:comparisons} cover a wide range of redshifts and sizes, a volume-limited sample of galaxies (or at least binning by redshift) is likely a much more appropriate comparison. While we leave detailed analysis to a further paper, we note that there is no strong change in the proportion of galaxies when binning by redshift intervals of $\Delta z=0.2$ out to $z=1.0$.
